%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Vectors and Matrices}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Scalars and vectors}
\label{sec:scalars_vectors}

The path of a car travelling from location $A$ to location $B$ is
characterized by two quantities or \emph{scalars}\index{scalars}: a
magnitude\index{magnitude}~(such as speed), and a direction~(from $A$
to $B$). Both of these two scalars are summarized by a
\emph{vector}\index{vector}.

\begin{figure}[!htpb]
\centering
\begin{tikzpicture}
[linedecorate/.style={->,>=stealth,very thick}]
% grids for the plane
\draw[step=1cm,lightgray,very thin] (-0.5,-0.5) grid (6.5,5.5);
% the rectangular axes
\draw[->,>=stealth,semithick] (-0.5,0) -- (6.5,0) node[right]{$x$};
\draw[->,>=stealth,semithick] (0,-0.5) -- (0,5.5) node[above]{$y$};
% vectors
\draw[linedecorate] (0,3) -- node[above left]{$\vecu$} (2,5);
\draw[linedecorate] (1.5,3.3) -- node[below right]{$\vecv$} (3.5,5.3);
\draw[linedecorate] (0.5,1.5) -- node[below right]{$\vecw$} (2.5,3.5);
\draw[linedecorate] (3.5,0.5) -- (5.5,0.5);
\draw[linedecorate] (5.7,0.7) -- (5.7,2.7);
\draw[linedecorate] (5.5,2.5) -- (3.5,2.5);
\draw[linedecorate] (3.3,2.3) -- (3.3,0.3);
\end{tikzpicture}
\caption{Vectors in the $x$-$y$ plane.}
\label{fig:vectors_matrices:plane_vectors}
\end{figure}

In the $x$-$y$ plane, we can visualize a vector as an arrow from point
$A$ to point $B$~(see
Figure~\ref{fig:vectors_matrices:plane_vectors}). The starting point
$A$ of the vector is called the \emph{tail}\index{vectors!tail} and
the terminal point $B$ is the \emph{head}.\index{vectors!head} Two
vectors are \emph{equivalent}\index{vectors!equivalent} if they have
the same magnitude and direction. The vectors $\vecu$, $\vecv$, and
$\vecw$ in Figure~\ref{fig:vectors_matrices:plane_vectors} are thus
all equivalent to each other.

To analyze vectors using algebra, we can think of a vector $\vecu$
as starting from the origin of the $x$-$y$ plane and having
$(u_1, u_2)$ as the coordinate for its head~(see
Figure~\ref{fig:specify_vector_head_coordinate}). So $\vecu$ is
completely determined by the coordinate of its head and we write
$\vecu = \langle u_1, u_2 \rangle$ as an algebraic representation
for $\vecu$.

\begin{figure}[!htpb]
\centering
\begin{tikzpicture}
% the rectangular axes
\draw[->,>=stealth,semithick] (-0.3,0) -- (4,0) node[right]{$x$};
\draw[->,>=stealth,semithick] (0,-0.3) -- (0,3.5) node[above]{$y$};
% vector described by its head coordinate
\draw[->,>=stealth,very thick] (0,0) -- node[above left]{$\vecu$} (3,2.5) node[below right]{$(u_1, u_2)$};
\end{tikzpicture}
\caption{A vector as specified by its head coordinate.}
\label{fig:specify_vector_head_coordinate}
\end{figure}

In case the tail of $\vecu$ is not the origin, we let $(x_1, y_1)$
be the coordinate of the tail and denote the coordinate of the head by
$(x_2, y_2)$. By subtracting coordinatewise, we obtain an equivalent
vector $\vecv$ that emanates from the origin of the $x$-$y$
plane. The coordinate of $\vecv$ is
\[
(x_2, y_2) - (x_1, y_1)
=
(x_2 - x_1,\; y_2 - y_1)
=
(u_1, u_2)
\]
and we do not distinguish between $\vecu$ and $\vecv$. Thus we
can transform $\vecu$ to be a vector emanating from the origin and
specified by the coordinate
%
\begin{equation}
\label{eq:component_form_vector}
\vecu
=
\langle x_2 - x_1,\; y_2 - y_1 \rangle
=
\langle u_1, u_2 \rangle.
\end{equation}
%
Equation~(\ref{eq:component_form_vector}) is called the
\emph{component form}\index{component form} of a vector, with $u_1$
and $u_2$ being the individual components. The vector
$\veczero = \langle 0, 0 \rangle$ is called the zero
vector.\index{zero vector}

\begin{figure}[!htpb]
\centering
\begin{tikzpicture}
% the rectangular axes
\draw[->,>=stealth,semithick] (-2.5,0) -- (5.7,0) node[right]{$x$};
\draw[->,>=stealth,semithick] (0,-1.5) -- (0,6.7) node[above]{$y$};
% ticks on horizontal axis
\foreach \x in {-2,-1,-1}
  \draw (\x cm,2pt) -- (\x cm,-2pt) node[anchor=north] {$\x$};
\foreach \x in {1,2,...,5}
  \draw (\x cm,2pt) -- (\x cm,-2pt) node[anchor=north] {$\x$};
% ticks on vertical axis
\foreach \y in {-1,-1}
  \draw (2pt,\y cm) -- (-2pt,\y cm) node[anchor=east] {$\y$};
\foreach \y in {1,2,...,6}
  \draw (2pt,\y cm) -- (-2pt,\y cm) node[anchor=east] {$\y$};
% vector u
\draw[->,>=stealth,very thick] (-2,-1) node[below]{$(-2,-1)$} -- node[above left]{$\vecu$} (1,3) node[above]{$(1,3)$};
% vector v
\draw[->,>=stealth,very thick] (1.5,1.5) node[below]{$(1.5,1.5)$} -- node[above left]{$\vecv$} (4.5,5.5) node[above]{$(4.5,5.5)$};
\end{tikzpicture}
\caption{Verify that vectors $\vecu$ and $\vecv$ are equivalent.}
\label{fig:vectors_matrices:verify_vectors_u_v}
\end{figure}

\begin{example}
Let the vector $\vecu$ be described by the directed line segment
from $(-2,-1)$ to $(1,3)$, and let $\vecv$ be the line segment from
$(1.5,1.5)$ to $(4.5,5.5)$, as shown in
Figure~\ref{fig:vectors_matrices:verify_vectors_u_v}. Show that
$\vecu$ and $\vecv$ are equivalent vectors.
\end{example}

\begin{proof}[Solution]
To show that $\vecu$ and $\vecv$ are equivalent, we need to show
that they have the same length and are in the same direction. The
length of $\vecu$ is equivalent to the length of the line segment
from $(-2,-1)$ to $(1,3)$:
%
\begin{align*}
\sqrt{(1 - (-2))^2 + (3 - (-1))^2}
&=
\sqrt{(1 + 2)^2 + (3 + 1)^2} \\
&=
\sqrt{9 + 16} \\
&=
5.
\end{align*}
%
This tells us that $\vecu$ has length 5. Similarly, the length of
$\vecv$ is
%
\begin{align*}
\sqrt{(4.5 - 1.5)^2 + (5.5 - 1.5)^2}
&=
\sqrt{3^2 + 4^2} \\
&=
\sqrt{9 + 16} \\
&=
5
\end{align*}
%
which is the same length as that of $\vecu$.

\begin{lstlisting}
sage: u = vector([1 - (-2), 3 - (-1)]); u
(3, 4)
sage: v = vector([4.5 - 1.5, 5.5 - 1.5]); v
(3.000...000, 4.000...000)
sage: u.norm()
5
sage: v.norm()
5.000...000
\end{lstlisting}

The slope of the line segment from $(-2,-1)$ to $(1,3)$ is
\[
\frac{3 - (-1)} {1 - (-2)}
=
\frac{4}{3}
\]
and the slope of the line segment from $(1.5,1.5)$ to $(4.5,5.5)$ is
\[
\frac{5.5 - 1.5}{4.5 - 1.5}
=
\frac{4}{3}
\]
which shows that these two line segments have the same direction, so
$\vecu$ and $\vecv$ point in the same direction.
%
\begin{lstlisting}
sage: (3 - (-1)) / (1 - (-2))
4/3
sage: (5.5 - 1.5) / (4.5 - 1.5)
1.333...333
sage: QQ((5.5 - 1.5) / (4.5 - 1.5))
4/3
\end{lstlisting}
%
Therefore, $\vecu$ and $\vecv$ are equivalent vectors.
\end{proof}

We denote the length of a vector as follows. Let $\vecu$ be a
vector whose starting point is $P = (x_1, y_1)$ and whose terminal
point is $Q = (x_2, y_2)$. Then the length or magnitude of $\vecu$
is denoted by $\vecvertl \vecu \vecvertr$ and given
by the equation
%
\begin{equation}
\label{eq:vectors_matrices:magnitude_two_dimensional_vector}
\begin{aligned}
\vecvertl \vecu \vecvertr
&=
\sqrt{(x_2 - x_1)^2 + (y_2 - y_1)^2} \\
&=
\sqrt{v_1^2 + v_2^2}.
\end{aligned}
\end{equation}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Add, subtract, and multiply vectors}
\index{vectors!arithmetic}
%% Relate vectors with complex numbers.

The left-hand side of Figure~\ref{fig:vectors_matrices:vector_sum}
illustrates the path of an object. Initially travelling in the
direction of vector $\vecu$, the object then changes its direction to
move in the direction of $\vecv$. The total displacement of the object
is represented as the vector $\vecu + \vecv = \vecw$. Furthermore, we
could first go in the direction of $\vecv$ and then in that of $\vecu$,
and hence obtain the same displacement $\vecv + \vecu = \vecw$, as
shown on the right-hand side of
Figure~\ref{fig:vectors_matrices:vector_sum}. This ``adding'' of
vectors is just one of many basic operations that can be performed on
vectors. In fact, the ``adding'' rule illustrated in
Figure~\ref{fig:vectors_matrices:vector_sum} is called the
\emph{parallelogram rule}\index{parallelogram rule} for vector
addition.

\begin{figure}[!htpb]
\centering
\begin{tikzpicture}
% left set of vectors
\node (start1) at (0,0) [circle,fill=blue,inner sep=1.5pt]{};
\node (middle1) at (3,0.1) [circle,inner sep=0.5pt]{};
\node (finish1) at (3.4,1.6) [circle,fill=blue,inner sep=1.5pt]{};
\draw[->,>=stealth,very thick] (start1) node[below left]{start}  -- node[below]{$\vecu$} (middle1);
\draw[->,>=stealth,very thick] (middle1) -- node[right]{$\vecv$} (finish1) node[above right]{finish};
\draw[->,>=stealth,very thick] (start1) -- node[above]{$\vecw$} (finish1);
% right set of vectors
\node (start2) at (6,0) [circle,fill=blue,inner sep=1.5pt]{};
\node (lowerMiddle) at (9,0.1) [circle,inner sep=0.5pt]{};
\node (upperMiddle) at (6.4,1.5) [circle,inner sep=0.5pt]{};
\node (finish2) at (9.4,1.6) [circle,fill=blue,inner sep=1.5pt]{};
\draw[->,>=stealth,very thick] (start2) node[below left]{start} -- node[below]{$\vecu$} (lowerMiddle);
\draw[->,>=stealth,very thick] (lowerMiddle) -- node[right]{$\vecv$} (finish2) node[above right]{finish};
\draw[->,>=stealth,shorten >=3pt,very thick] (start2) -- node[above]{$\vecw$} (finish2);
\draw[->,>=stealth,very thick,dash pattern=on 4pt off 4pt] (start2) -- node[left]{$\vecv$} (upperMiddle);
\draw[->,>=stealth,very thick,dash pattern=on 4pt off 4pt] (upperMiddle) -- node[above]{$\vecu$} (finish2);
\end{tikzpicture}
\caption{The vector sum $\vecu + \vecv = \vecw = \vecv + \vecu$.}
\label{fig:vectors_matrices:vector_sum}
\end{figure}

\begin{definition}
\label{def:vectors_matrices:vector_operations}
Let $\vecu = \langle u_1, u_2 \rangle$ and
$\vecv = \langle v_1, v_2 \rangle$ be vectors and suppose
$c$ is a scalar. Then we have the following operations on vectors:
\begin{enumerate}
\item The \emph{vector sum} of $\vecu$ and
  $\vecv$ is $\vecu + \vecv = \langle u_1 + v_1,\; u_2 + v_2 \rangle$.

\item The \emph{negative} of $\vecu$ is
  $-\vecu = \langle -u_1, -u_2 \rangle$.

\item The \emph{vector difference} of $\vecu$ and
  $\vecv$ is $\vecu - \vecv =
  \langle u_1, u_2 \rangle - \langle v_1, v_2 \rangle = \langle u_1 -
  v_1,\; u_2 - v_2 \rangle$.

\item The \emph{scalar multiple} of $c$ and $\vecu$ is
  $c\vecu = \langle c u_1, c u_2 \rangle$.
\end{enumerate}
\end{definition}

The negative of a vector $\vecu$ is a vector that has the same
length as $\vecu$, but goes in the opposite direction. One
exception is the zero vector $\veczero$, whose negative is itself
because $-\veczero = -\langle 0,0 \rangle = \langle -0, -0 \rangle =
\langle 0,0 \rangle = \veczero$. More generally, scalar
multiplication has the effect of scaling a vector. This may result in
expanding the vector, contracting it, producing the negative of the
vector, or expanding/contracting the negative of the vector. These
various possibilities are shown in
Figure~\ref{fig:scalar_multiply_negative}.

\begin{figure}[!htpb]
\centering
\begin{tikzpicture}
\draw[->,>=stealth,very thick] (0,0) -- (1,1) node[above]{$\vecu$};
\draw[<-,>=stealth,very thick] (1,0) -- (2,1) node[above]{$-\vecu$};
\draw[->,>=stealth,very thick] (2,0) -- (2.5,0.5) node[above right]{$\frac{1}{2}\vecu$};
\draw[->,>=stealth,very thick] (3,0) -- (5,2) node[above]{$2\vecu$};
\draw[<-,>=stealth,very thick] (4,0) -- (5.5,1.5) node[above right]{$-\frac{3}{4}\vecu$};
\end{tikzpicture}
\caption{Scalar multiplication and negative of a vector.}
\label{fig:scalar_multiply_negative}
\end{figure}

Using the parallelogram rule for vector addition, we can similarly
illustrate vector difference. The left-hand side of
Figure~\ref{fig:vectors_matrices:vector_difference_parallelogram_triangle} shows the
parallelogram rule for vector subtraction, while the right-hand side shows
the triangle rule for vector subtraction. Thus to determine the
difference of two vectors $\vecu$ and $\vecv$, we first align the
vectors so that their initial points coincide. Then the difference
$\vecu - \vecv$ is the vector that starts from the head of
$\vecv$ and ends at the terminal point of $\vecu$.

\begin{figure}[!htpb]
\centering
\begin{tikzpicture}
% left-hand figure: vector difference by parallelogram rule
\node (origin) at (0,0) [circle,fill=blue,inner sep=1.5pt]{};
\node (destin) at (-1.7,2) [circle,fill=blue,inner sep=1.5pt]{};
\draw[->,>=stealth,very thick] (origin) -- node[below]{$\vecv$} (2.5,0);
\draw[->,>=stealth,very thick] (origin) -- node[below]{$-\vecv$} (-2.5,0);
\draw[->,>=stealth,very thick] (origin) -- node[right]{$\vecu$} (0.8,2);
\draw[->,>=stealth,very thick,dash pattern=on 4pt off 4pt] (-2.5,0) -- node[left]{$\vecu$} (destin);
\draw[->,>=stealth,very thick,dash pattern=on 4pt off 4pt] (0.5,2) -- node[above]{$-\vecv$} (destin);
\draw[->,>=stealth,very thick] (origin) -- node[right]{$\vecu - \vecv$} (destin);
% right-hand figure: vector difference
\draw[->,>=stealth,very thick] (4.5,0) -- node[left]{$\vecu$} (5.3,2);
\draw[->,>=stealth,very thick] (4.5,0) -- node[below]{$\vecv$} (7,0);
\draw[->,>=stealth,very thick] (7,0) -- node[right]{$\vecu - \vecv$} (5.3,2);
\end{tikzpicture}
\caption{Vector difference by the parallelogram rule~(left) and by the
triangle rule~(right).}
\label{fig:vectors_matrices:vector_difference_parallelogram_triangle}
\end{figure}

Vector operations share many of the familiar properties of operations
on real numbers. Given two real numbers, for example, it does not
matter in which order we add
them. Problem~\ref{prob:vectors_matrices:field_laws_for_vectors} lists
a number of vector rules that are similar to those obeyed by real numbers.

\begin{example}
Using vector rules from
problem~\ref{prob:vectors_matrices:field_laws_for_vectors}, find a
vector $\vecx$ such that $5\vecx + 7\vecu = 4\vecv$.
\end{example}

\begin{proof}[Solution]
First, subtract $7\vecu$ from both sides of the equal sign to get
$5\vecx + (7\vecu - 7\vecu) = 4\vecv - 7\vecu$, which can be
simplified to $5\vecx = 4\vecv - 7\vecu$. Now divide both sides of
the last equation by $5$ and we have
$\frac{1}{5} (5\vecx) = \frac{1}{5} (4\vecv - 7\vecu)$. Simplify the
last equation to
$\vecx = \frac{1}{5} (4\vecv) + \frac{1}{5}(-7\vecu)$. So the required
vector is $\vecx = \frac{4}{5}\vecv - \frac{7}{5}\vecu$.
\end{proof}

Any nonzero vector $\vecv$ can be transformed into a vector $\vecu$ of
magnitude $1$. The equation to effect this transformation is
%
\begin{equation}
\label{eq:vectors_matrices:two_dimensional_unit_vector}
\vecu
=
\frac {\vecv} {\vecvertl \vecv \vecvertr}
=
\vecv
\frac {1} {\vecvertl \vecv \vecvertr}.
\end{equation}
%
The process of using
equation~(\ref{eq:vectors_matrices:two_dimensional_unit_vector}) to
transform a nonzero vector into a vector of magnitude $1$ is called
\emph{normalizing}\index{vectors!normalization} a vector. For example,
given the vector $\vecu = \langle 1, 2 \rangle$, we can use
equation~(\ref{eq:vectors_matrices:two_dimensional_unit_vector}) to
find the normalized form of $\vecu$. Since
$\vecvertl \vecu \vecvertr = \sqrt{5}$ by
equation~(\ref{eq:vectors_matrices:magnitude_two_dimensional_vector}),
so $\vecu$ is normalized as
%
\begin{align*}
\frac {\vecu} {\vecvertl \vecu \vecvertr}
&=
\frac{1}{\sqrt{5}} \langle 1, 2 \rangle \\[4pt]
&=
\frac{\sqrt{5}}{5} \langle 1, 2 \rangle.
\end{align*}

\begin{lstlisting}
sage: u = vector([1, 2])
sage: u / u.norm()
(1/5*sqrt(5), 2/5*sqrt(5))
\end{lstlisting}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Applications to geometry}
\label{subsec:2D_vectors_apply_geometry}

A line in the $x$-$y$ plane is usually described by the linear
equation
\[
y = mx + c
\]
where $m \neq 0$ is the slope of the line and $c$ is a constant. The
same line can be represented in \emph{vector form}\index{vector form}
as
%
\begin{equation}
\label{eq:vector_form_line}
\langle x,\, mx + c \rangle
=
\langle 0, c \rangle + x \langle 1, m \rangle.
\end{equation}
%
The vector $\langle 0, c \rangle$ points to the line, while the vector
$\langle 1, m \rangle$ is in the same direction as the line.

\begin{example}
Express the line $2x + 3y = 4$ in vector form.
\end{example}

\begin{proof}[Solution]
Solving the equation $2x + 3y = 4$ for $y$ to get
\[
y = \frac{4}{3} - \frac{2}{3} x.
\]
Using equation~(\ref{eq:vector_form_line}), we have
\[
\langle 0,\, 4/3 \rangle
+ x \langle 1,\, -2/3 \rangle
\]
which is a representation of the line $2x + 3y = 4$ in vector form.
\end{proof}

In general, a line $L$ has many different vector representations. If
$(u_1, u_2)$ is a point on $L$, then $\vecu = \langle u_1, u_2
\rangle$ is a vector pointing to $L$. Furthermore, let $\vecv =
\langle v_1, v_2 \rangle$ be a vector in the same direction as
$L$. Then the line $L$ can be represented in vector form as
\[
\langle x, y \rangle
=
\vecu + t \vecv
\]
with $t$ being a parameter.

Two nonzero vectors $\vecu$ and $\vecv$ are said to be
\emph{linearly dependent}\index{linearly dependent} if they can be
expressed as a linear combination that equals the zero vector, i.e. we
can find two nonzero real numbers $m,n$ such that
$m\vecu + n\vecv = \veczero$. From this definition, it follows that
two vectors are parallel if and only if they are linearly
dependent. In other words, if two vectors are parallel then they are
linearly dependent. The converse is also true: if two vectors are
linearly dependent, then they are parallel.
